---
title: "R Notebook"
output: html_notebook
---



```{r}
library(ggplot2)

# Loading the data
DO <- read.csv("Berthomieux2012/mainTextData/msb201270-df3A.txt",
               skip = 52,header = FALSE)

names(DO) <- c("time","DO")
fakeColors <- factor(sample(1:5,length(DO[,1]),replace=TRUE))
```


Plotting scatterplots *geom_point()*


```{r}
# Creating a simple scatter plot
p <- ggplot(DO, aes(time,DO)) + geom_point()
plot(p)
```


Adjusting the color of the points *colour*

```{r}
ggplot(DO, aes(time,DO)) + geom_point(colour="green")
```


Adjusting the size of the points *size*

```{r}
ggplot(DO, aes(time,DO)) + geom_point(size=10)
```

Changing point shapes *shape*

```{r}
ggplot(DO,aes(time, DO)) + geom_point(shape=5)
```

Adding color levels *colour=factor*

Colour parameter can be assigned inside aes() and geom_point().

```{r}
ggplot(DO, aes(time,DO)) + geom_point(colour=fakeColors)
```


Create a gradient of colors *scale_color_gradient()*

```{r}
ggplot(DO, aes(time, DO, colour=DO)) + geom_point() + scale_color_gradient(low="red",high = "yellow")
```

Creating non filled points *solid=FALSE*

```{r}
ggplot(DO, aes(time,DO)) + geom_point(shape=1) + scale_shape(solid = FALSE) 
```

Fitting a smooth line model to a scatter plot *geom_smooth()*

```{r}
ggplot(DO, aes(time,DO)) + geom_point(colour="green") + geom_smooth()
```

Mixing point styles *geom_point()*

```{r}
p <- ggplot(DO, aes(time, DO)) +   geom_point(size=8) + geom_point(colour="green",size=2)
plot(p)
```




